---
title: "218Y_Beck_A4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, progress_bar = F)
```

```{r}
library(remotes)
library(tidyverse)
library(sf)
library(leaflet)
library(mapboxapi)
library(tigris)
library(esri2sf)
library(raster)
library(stars)
library(plotly)
library(tidycensus)
library(censusapi)
library(mapview)

Sys.setenv(CENSUS_KEY="c18a7f6e5cbbd1e7f0d3047d8b674cbaad89bcb3")
```

<h3> Part 1: Measuring Sea Level Rise Over Time </h3>

<br> Map of Redwood City's Boundaries <br>
```{r}
ca_cbgs <- block_groups("CA", cb = T, progress_bar = F)

rdwd_boundary <- places("CA", cb = T, progress_bar = F) %>% 
  filter(NAME == "Redwood City")

rdwd_cbgs_map <- 
  ca_cbgs %>% 
  st_centroid() %>% 
  .[rdwd_boundary, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_cbgs %>% select(GEOID)) %>% 
  st_as_sf()

##Map Redwood City 

leaflet() %>% 
  addMapboxTiles(
    style_id = "streets-v11",
    username = "mapbox"
  ) %>% 
  addPolygons(
    data = rdwd_cbgs_map,
    fill = F
  )
```

```{r}

slr <- 25
rp <- 20

path <- paste0("/Volumes/GoogleDrive-110275654074378237038/Shared drives/SFBI/Data Library/OCOF/san_mateo_flooding_slr", str_pad(slr, 3, "left","0" ),"/flooding/v2.1/county_san_mateo_flddepth_slr",str_pad(slr, 3, "left", "0"),"_w",str_pad(rp, 3, "left", "0"),".tif")

test_flood <- raster(path)

ca_cbgs <- block_groups("CA", cb = T, progress_bar = F)

rdwd_boundary <- places("CA", cb = T, progress_bar = F) %>%
  filter(NAME == "Redwood City") 


##filter by cbgs with greatest AWater 
rdwd_cbgs <- 
  ca_cbgs %>% 
  st_centroid() %>% 
  .[rdwd_boundary, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_cbgs %>% select(GEOID)) %>% 
  filter(GEOID %in% c("060816103021", "060816103032", "060816103034")) %>%
  st_as_sf()


test_flood_rdwd <- test_flood %>%
  crop(
    rdwd_cbgs %>%
      st_transform(26910) %>%
      st_bbox()
  )

test_flood_rdwd_raster <- test_flood %>%
  crop(
    rdwd_cbgs %>%
    filter(GEOID == "060816103032") %>%
      st_transform(26910) %>%
      st_bbox()
  )

plot(test_flood_rdwd)

print("This is the flood map of Redwood City")


flood_pal <- colorNumeric(
  palette = "Blues",
  domain = values(test_flood_rdwd),
  na.color = "transparent"
)

leaflet() %>%
  addMapboxTiles(
    style_id = "satellite-streets-v11",
    username = "mapbox",
    options = tileOptions(opacity = 0.5)
  ) %>% 
  addRasterImage(
    test_flood_rdwd_raster,
    colors = flood_pal
  ) %>% 
  addLegend(
    pal = flood_pal,
    values = values(test_flood_rdwd_raster),
    title = "Flood depth, cm"
  )
```
<br> Here we can see an image of the Redwood City flood basin. Clearly this area is due to experience damaging effects from sea level rise <br> 

```{r}
##pass through Redwood City
# for(slr in c("000","025","050")){ ##sea level rise options
# 
#   for(rp in c("001","020","100")){ ##time periods
# 
#     print(paste0("SLR",slr,"_RP",rp))
# 
#     path <-  paste0("/Volumes/GoogleDrive-110275654074378237038/Shared drives/SFBI/Data Library/OCOF/san_mateo_flooding_slr", str_pad(slr, 3, "left","0" ),"/flooding/v2.1/county_san_mateo_flddepth_slr",str_pad(slr, 3, "left", "0"),"_w",str_pad(rp, 3, "left", "0"),".tif")
# 
#       # paste0("san_mateo_flooding_slr",slr,"/flooding/v2.1/county_san_mateo_flddepth_slr",slr,"_w",rp,".tif")
# 
#     flood <- raster(path) %>%
#       crop(
#         rdwd_cbgs %>%
#           st_transform(26910) %>%
#           st_bbox()
#       )
# 
#     writeRaster(flood, paste0("flood/SLR",slr,"_RP",rp,"_rdwd_flood.tif"), overwrite = T)
# 
#   }
# }

```

```{r}
# osm_bldg <- st_read("/Volumes/GoogleDrive-110275654074378237038/Shared drives/SFBI/Data Library/OSM/gis_osm_buildings_a_free_1.shp")
# 
# rdwd_bldg <- osm_bldg[rdwd_boundary, ]
```


```{r}
##load in osm data and create rdwd building 
# osm_bldg <- readRDS("osm_bldg.rds")
# 
# rdwd_cbgs <- rdwd_cbgs %>%
# st_transform(4326)
# 
# rdwd_cbg_bldg <- osm_bldg[rdwd_cbgs, ]
# 
# saveRDS(rdwd_cbg_bldg, "rdwd_cbg_bldg.rds")

rdwd_cbg_bldg <- readRDS("rdwd_cbg_bldg.rds")
```

<h3> Part 2:  Building Exposure in Redwood City </h3> 

```{r}
##Flood potential 
flood_max <- 
  raster("flood/SLR050_RP100_rdwd_flood.tif")

flood_max_extent <- 
  flood_max %>% 
  st_as_stars() %>% 
  mutate(SLR050_RP100_rdwd_flood = ifelse(
    !is.na(SLR050_RP100_rdwd_flood),
    1,
    NA
  )) %>% 
  st_as_sf(merge = T) %>% 
  st_set_crs(26910) %>% 
  st_make_valid() %>% 
  st_transform(4326)

rdwd_bldg_flooded_max <-
  rdwd_cbg_bldg %>% 
  st_transform(4326) %>% 
  .[flood_max_extent,]

#saveRDS(rdwd_bldg_flooded_max,"rdwd_bldg_flooded_max.rds")
 
# rdwd_bldg_exposure <- NULL
# 
# for(slr in c("000","025","050")){
#   
#   for(rp in c("001","020","100")){
#     
#     print(paste0("SLR",slr,"_RP",rp))
#     
#     flood <- raster( paste0("flood/SLR",slr,"_RP",rp,"_rdwd_flood.tif"))
#     
#     flood_extent <- 
#       (flood > -Inf) %>% 
#       st_as_stars() %>% 
#       st_as_sf(merge = T) %>% 
#       st_set_crs(26910) %>% 
#       st_make_valid() %>% 
#       st_transform(4326)
#     
#     rdwd_bldg_flooded <-
#       rdwd_bldg_flooded_max[flood_extent,] %>% 
#       st_transform(26910)
#     
#     flood_crop <-
#       crop(flood, rdwd_bldg_flooded)
#     
#     flood_crop[is.na(flood_crop)] <- 0
#     
#     temp <-
#       raster::extract(
#         flood_crop,
#         rdwd_bldg_flooded,
#         fun = mean
#       ) %>% 
#       as.data.frame() %>% 
#       rename(avg_depth = V1) %>% 
#       cbind(
#         rdwd_bldg_flooded %>% 
#           st_drop_geometry() %>% 
#           dplyr::select(osm_id)
#       ) %>% 
#       mutate(
#         SLR = slr,
#         RP = rp
#       )
#     
#     rdwd_bldg_exposure <- 
#       rdwd_bldg_exposure %>% 
#       rbind(temp)
#     
#   }
# }
# saveRDS(rdwd_bldg_exposure,"rdwd_bldg_exposure.rds")

```


```{r}
##load in building exposure data
rdwd_bldg_exposure <- readRDS("rdwd_bldg_exposure.rds") ##turn average depth into values currency 
```

```{r}
##Load in EMFAC data
emfac <- read_csv("EMFACFutureSM.csv", skip = 8)

future_vehicle_count <- 
  emfac %>%
  dplyr::select(
    year = `Calendar Year`,
    count = Population
  ) %>%
  group_by(year) %>%
  summarize(count = sum(count)) %>%
  mutate(
    perc = count/min(count)
  )

```


```{r}
##get census data for number of vehicles in a block group 
rdwd_cbg <- block_groups("CA","San Mateo", cb = F, progress_bar = F, year = 2019) %>% 
  filter(GEOID %in% c("060816103021", "060816103032", "060816103034")) %>% 
  st_transform(4326)

rdwd_bldgs <- readRDS("rdwd_cbg_bldg.rds") %>% 
  .[rdwd_cbg,]

# rdwd_blocks <-
#    blocks("CA","San Mateo", progress_bar = F, year = 2020)
# saveRDS(rdwd_blocks, "rdwd_blocks.rds")

rdwd_blocks <-
  readRDS("rdwd_blocks.rds") %>% 
  st_transform(4326) %>% 
  .[rdwd_bldgs,]

acs_vars_2019_5yr <- readRDS("acs_vars_2019_5yr.rds")

rdwd_blocks_pop <-
  getCensus(
    name = "dec/pl",
    vintage = 2020,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "P1_001N" ##population data
  ) %>% 
  transmute(
    GEOID20 = paste0(state,county,tract,block),
    pop = P1_001N
  ) %>% 
  filter(GEOID20 %in% rdwd_blocks$GEOID20)

rdwd_bg_vehicle <-
  getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "block group:*",
    regionin = "state:06+county:081",
    vars = "group(B25044)" ##vehicle ownership 
  ) %>% 
  mutate(
    cbg =
      paste0(state,county,tract,block_group)
  ) %>% 
  dplyr::select(!c(GEO_ID,state,county,tract,block_group,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_5yr %>% 
      dplyr::select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  dplyr::select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,"tenure", "vehicle"),
    sep = "!!"
  ) %>% 
  filter(!is.na(vehicle)) %>% 
  filter(cbg %in% rdwd_cbg$GEOID) ##filter by the cbg group 

rdwd_veh_zero_one <- rdwd_bg_vehicle  %>%
  group_by(vehicle) %>%
  summarize(estimate = sum(estimate))


##Find total number of vehicles in our study cbgs
rdwd_bg_vehicle_total <- rdwd_bg_vehicle %>% 
  filter(vehicle != "No vehicle available") %>% 
  mutate(
    vehicle = substr(vehicle, 1,1) %>% as.numeric(),
    vehicle_count = vehicle * estimate
  ) %>% 
  group_by(cbg) %>% 
  summarize(vehicle_count = sum(vehicle_count))


```
Out of total households, 81 of them did not have any vehicles available in 2020. 919 of them had 1 vehicle. 

```{r}
##Look to 2020-2050
rdwd_veh_projection <- rdwd_bg_vehicle_total %>%
  summarize(vehicle_count = sum(vehicle_count)) %>% ##add up all the vehicle counts in our cbg area 
  cbind(future_vehicle_count) %>%
  select(-count) %>%
  mutate(
   vehicle_count = vehicle_count *perc
  )  %>%
  mutate(
    no_vehicles = 81 * perc
  ) %>%
  mutate(
    one_vehicle = 919* perc
  ) %>%
  select(-perc)

rdwd_veh_projection
```
<br> Assuming each city in San Mateo is increasing vehicle count by the same amount each decade, we can use EMFAC data to determine how many vehicles there will be in our flood risk zone between 2020-2050. In 2020 there is expected to be 5645 cars, in 2030 there is expected to be 7997.906, in 2030 there is expected be 9487.590, and in 2050 there is expected to be 10106.098 vehicles. Households with no vehicles in our study area is also projected to increase by the same percentage and same with households with one vehicle. When determining flood risk for these vehicles, we need to remember that we are using household flood risk as our test, so it will look like there is little to no risk for those households with 1/0 vehicles. Obviously, this is not the case, however, this particular model is looking at vehicle-related flood damage and thus cannot capture accurately the damage incurred by households.  <br>
<br> 


```{r}
rdwd_block_veh_per_bldg <-
  rdwd_bldgs %>% 
  filter(type %in% c("house", "apartments", "detached", "residential", NA)) %>% # any non-residential buildings?
  select(osm_id) %>% # unique ID for each building
  st_centroid() %>% 
  st_join(rdwd_blocks %>% select(GEOID20)) %>% # block shapes
  st_join(rdwd_cbg %>% select(cbg = GEOID)) %>% # cbg shapes
  st_drop_geometry() %>% 
  group_by(GEOID20, cbg) %>% # "assuming population is distributed evenly across buildings in a block"
  summarize(bldg_count = n()) %>% # how to get counts? add up all the unique building codes
  left_join(rdwd_blocks_pop) %>% # census dataset
  left_join(rdwd_bg_vehicle_total) %>% # census dataset
  group_by(cbg) %>% # "and vehicles are distributed evenly across population"
  mutate(
    veh_per_person = vehicle_count/sum(pop),
    ppl_per_bldg = pop/bldg_count,
    veh_per_bldg = veh_per_person*ppl_per_bldg # fractional result ok
  )

#saveRDS(rdwd_block_veh_per_bldg, "rdwd_block_veh_per_bldg.rds")

# dataset with individual buildings, vehicle counts for each building
rdwd_veh_per_bldg <- rdwd_bldgs %>% 
  filter(!is.na(type)) %>% 
  select(osm_id) %>% 
  st_centroid() %>% 
  st_join(rdwd_blocks %>% select(GEOID20)) %>% 
  left_join(rdwd_block_veh_per_bldg %>% select(GEOID20, veh_per_bldg))
```


<h3> Part 4: Vulnerability Data </h3>
```{r}
vulnerability <- data.frame(
  depth = c(0,0.5,1,2,3,4,5,6,7,8,9,10),
  perc_damage = c(
    0,
    0.076,
    0.28,
    0.462,
    0.622,
    0.76,
    0.876,
    0.97,
    1,
    1,
    1,
    1
  )
)


rdwd_vehicle_exposure <- 
  readRDS("rdwd_bldg_exposure.rds") %>% 
  mutate(
    avg_depth = avg_depth*0.0328084 # cm to ft
  )

rdwd_vehicle_perc_damage <- 
  approx(
    x = vulnerability$depth,
    y = vulnerability$perc_damage,
    xout = rdwd_vehicle_exposure$avg_depth
  ) %>% 
  .[2] %>% 
  as.data.frame() %>% 
  rename(perc_damage = y) %>% 
  cbind(rdwd_vehicle_exposure)

##saveRDS(rdwd_vehicle_perc_damage,"rdwd_vehicle_perc_damage.rds")
```
<br> we are able to use building damage as a proxy for vehicle damage with the assumption that street level or basement flooding would affect cars as well as houses. We swapped the average depth of exposure to 0.5 ft instead of the -2 in the textbook example. <br> 
```{r}
##Plot the damage

##damage based on individual buildings
##Plot the vehicle damage
rdwd_vehicle_perc_damage_plot <- 
  expand.grid(
    osm_id = unique(rdwd_vehicle_perc_damage$osm_id),
    SLR = unique(rdwd_vehicle_perc_damage$SLR),
    RP = unique(rdwd_vehicle_perc_damage$RP)
  ) %>% 
  left_join(rdwd_vehicle_perc_damage) %>% 
  mutate(
    avg_depth = ifelse(
      is.na(avg_depth),
      0,
      avg_depth
    )
  )

rdwd_plot <- 
  plot_ly() %>% 
  add_trace(
    data = 
      rdwd_vehicle_perc_damage_plot %>% 
        filter(RP == "100") %>% 
        mutate(SLR = SLR %>% as.numeric()),
    x = ~avg_depth,
    y = ~perc_damage,
    frame = ~SLR,
    type = 'scatter',
    mode = 'markers',
    marker = list(
      color = 'rgba(17, 157, 255, 0.01)',
      size = 15
    ),
    showlegend = F
  ) %>% 
  add_trace(
    data = vulnerability,
    x = ~depth,
    y = ~perc_damage,
    type = 'scatter',
    mode = 'markers',
    marker = list(
      color = 'rgb(0,0,0)'
    ),
    showlegend = F
  ) %>% 
  layout(
    xaxis = list(
      title = "Average Flood Depth",
      zeroline = FALSE
    ),
    yaxis = list(
      title = "Percent Damage"
    ),
    title = "Redwood City vehicle damage during<br>100-year storm, by base sea level rise"
  ) %>% 
  config(displayModeBar = F)

rdwd_plot
```
<br>

The more flood depth, the greater the vehicle flood damage as can be seen in this plot. 

<h3> Part 5: Average Anualized Cost of Floods </h3> 
<br> Risk Estimation <br>
```{r}
projection <- "+proj=utm +zone=10 +ellps=GRS80 +datum=NAD83 +units=ft +no_defs"

rdwd_vehicle_flooded_max <- 
  readRDS("rdwd_bldg_flooded_max.rds") %>% 
  st_transform(projection) %>% 
  mutate(
    area = st_area(.) %>% as.numeric()
  )

rdwd_vehicle_damage <-
  rdwd_vehicle_perc_damage %>% 
  left_join(
    rdwd_veh_per_bldg %>% 
      st_drop_geometry() %>% 
      select(osm_id, veh_per_bldg)
  ) %>% 
  filter(!is.na(veh_per_bldg)) %>%
  mutate(
    damage = veh_per_bldg * 0.119 * 24112* perc_damage 
  ) %>% 
  select(osm_id, SLR, RP, damage)

#head(rdwd_vehicle_damage) ##$$ damage for each vehicle, for each event

```
<br> In order to compute the damage function we determined that damage is equal to the average cost of a vehicle (determined by resale value computed by the New York Times https://www.nytimes.com/2021/03/25/business/car-paint-job-resale-value.html) and mulitply that by perc_damage and the number of cars that will be flooded and the number of vehicles per building.
Per the research, 88% of vehicles would be moved to higher ground if given 12 hours notice of a flood, therefore we need to factor that into expected damage. 
<br>
This table shows us the damage for vehicle exposure in redwood city. Clearly OSM 233058123 has the highest $ damage risk. 
<br>
Here we have the  “$ damage” for each vehicle, for each event, we can now combine each trio of storm events together (for each of 5 levels of sea level rise). The result will be an “average annualized loss” for each vehicle for 5 different hypothetical years, each of which has a different base sea level rise.


```{r}
rdwd_vehicle_aal_by_slr <-
  rdwd_vehicle_damage %>% 
  pivot_wider(
    names_from = RP,
    values_from = damage
  ) %>% 
  replace(is.na(.), 0) %>% 
  mutate(
    damage = 
      0.95*(`001`+`020`)/2 + 
      0.04*(`020`+`100`)/2 + 
      0.01*(`100`)
  ) %>% 
  select(osm_id, SLR, damage)

```
<br> For each of our study OSM areas we have a dollar value assigned to the damage. Clearly OSM 233058123 has the highest cost related to flood risk. 

<br> 
 Next, we will consider, for any given year (we’ll bound our analysis to the 2020-2050 range), the likelihood of sea level rise being some amount or greater. Intuitively for the current year, the current sea level rise is what it is, and the chances of any greater amount of sea level rise before the end of the year are effectively 0. 10 years from now, the distribution of probabilities will be something, based on climate models.
<br>
```{r}
rcp45 <- read_csv("https://raw.githubusercontent.com/stanfordfuturebay/stanfordfuturebay.github.io/master/advanced/rcp45_sanfrancisco.csv")

rcp45
```

```{r}
rdwd_vehicle_aal_by_year <- 
  rdwd_vehicle_aal_by_slr %>% 
  left_join(
    rcp45 %>% 
      mutate(
        SLR = str_pad(SLR, 3 , "left", "0")
      ) %>% 
      select(
        SLR,
        `2020`,
        `2030`,
        `2040`,
        `2050`
      )
  ) %>% 
  pivot_longer(
    `2020`:`2050`,
    names_to = "year",
    values_to = "occurrence"
  ) %>% 
  pivot_longer(
    c(damage,occurrence),
    names_to = "key",
    values_to = "value"
  ) %>% 
  pivot_wider(
    names_from = c("key","SLR"),
    values_from = value
  ) %>% 
  replace(is.na(.), 0) %>% 
  mutate(
    damage = 
      occurrence_000 * (damage_000 + damage_025)/2 + 
      occurrence_025 * (damage_025 + damage_050)/2 + 
      occurrence_050 * (damage_050)
  ) %>% 
  select(osm_id, year, damage)

#head(rdwd_vehicle_aal_by_year)
##bldg footprint is the proxy for the vehicles at street level and garage 
```
<br> Now we have projected flood risk between 2020-2050 and its associated $ damage. It seems like 2020 and 2030 are going to be the most costly years with their associated osm_id's. <br>

```{r}
##Final AALs in Our Study Period

total_aal <- rdwd_vehicle_aal_by_year %>%
  group_by(year) %>%
  summarize(damage = sum(damage))

total_aal
```
<br> The $ damage function per year during our study interval can be seen in the table above. The damage increases exponentially each decade. <br>

```{r}
rdwd_vehicle_aal_by_year_map <-
  rdwd_vehicle_aal_by_year %>% 
  pivot_wider(
    names_from = year,
    values_from = damage
  ) %>% 
  mutate(
    change = `2050`-`2020`
  ) %>% 
  left_join(
    rdwd_vehicle_flooded_max %>%
      select(osm_id)
  ) %>% 
  st_as_sf() %>% 
  st_transform(4326)

aal_pal <- colorNumeric(
  palette = "Reds",
  domain = c(0,rdwd_vehicle_aal_by_year_map$`2050`)
)

rdwd_vehicle_aal_by_year_map %>% 
  leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolygons(
    fillColor = ~aal_pal(`2020`),
    color = "gray",
    fillOpacity = 1,
    opacity = 1,
    weight = 0.25,
    highlightOptions = highlightOptions(
      color = "white",
      weight = 2
    ),
    label = ~paste0("$",prettyNum(signif(`2020`,2),",")," average annualized loss in 2020"),
    group = "2020"
  ) %>% 
  addPolygons(
    fillColor = ~aal_pal(`2050`),
    color = "gray",
    fillOpacity = 1,
    opacity = 1,
    weight = 0.25,
    highlightOptions = highlightOptions(
      color = "white",
      weight = 2
    ),
    label = ~paste0("$",prettyNum(signif(`2050`,2),",")," average annualized loss in 2050"),
    group = "2050"
  ) %>% 
  addPolygons(
    fillColor = ~aal_pal(change),
    color = "gray",
    fillOpacity = 1,
    opacity = 1,
    weight = 0.25,
    highlightOptions = highlightOptions(
      color = "white",
      weight = 2
    ),
    label = ~paste0("$",prettyNum(signif(change,2),",")," change in average annualized loss from 2020 to 2050"),
    group = "Change"
  ) %>% 
  addLegend(
    pal = aal_pal,
    values = ~`2050`,
    title = "AAL"
  ) %>% 
  addLayersControl(
    baseGroups = c("2020","2050","Change"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% 
  showGroup("2050")
```
<br> 
When toggling between 2020 and 2050, we can see that there is always a flood risk in this zone that is only exarcerbated over time with some buildings and vehicles becoming more exposed in the inlet close to the Bayshore Freeway (some being bright orange/red). The peninsula with Saginew Dr in the center seems to be most at risk in 2050 with the greatest change between 2020 and 2050. The Redwood Shores Lagoon neighborhood is at serious risk of flood damage though none of the buildings seem to have as urgent or dire a flood risk as the Maple st neighborhood.  Luckily, there does seem to be a fair amount of buffer zone between the water and housing. This may be because we did not include industrial use buildings and if we did there would be greater red zones. 

Obviously there are a lot of assumptions in our study and a lot of projection, but clearly Redwood City needs to strengthen its climate change/sea level rise mitigation plans.<br> 

<br> Redwood AAL <br>
```{r}
rdwd_bg_aal <-
  rdwd_vehicle_aal_by_year %>% 
  pivot_wider(
    names_from = year,
    values_from = damage
  ) %>% 
  mutate(
    aal = (`2020`*5 + `2030`*10 + `2040`*10 + `2050`*5)/30
  ) %>% 
  left_join(
    rdwd_vehicle_flooded_max %>%
      select(osm_id) %>% 
      st_centroid()
  ) %>% 
  st_as_sf() %>% 
  st_transform(4269) %>% 
  st_join(rdwd_cbgs) %>% 
  st_set_geometry(NULL) %>% 
  group_by(GEOID) %>% 
  summarize(
    aal = sum(aal),
    count = n()
  ) %>% 
  left_join(rdwd_cbgs) %>% 
  st_as_sf()

final_table <-
  rdwd_bg_aal %>%
  select(GEOID, aal) %>%
  st_drop_geometry()

print(final_table)

sum(final_table$aal) %>% prettyNum(",") %>% paste0("$",.)
```
<br> Across our three block groups, these are the averaage anual loss in dollars values. The block group with the highest AAL is the Redwood Shores Lagoon area, which is consistent with the previous map. <br> 
```{r}
aal_pal <- colorNumeric(
  palette = "Reds",
  domain = rdwd_bg_aal$aal
)

rdwd_bg_aal %>% 
  leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolygons(
    fillColor = ~aal_pal(aal),
    color = "gray",
    fillOpacity = 0.5,
    opacity = 1,
    weight = 0.25,
    highlightOptions = highlightOptions(
      color = "white",
      weight = 2
    ),
    label = ~paste0("$",prettyNum(signif(aal,2),",")," average annualized loss across ", count, " buildings, 2020-2050")
  ) %>% 
  addLegend(
    pal = aal_pal,
    values = ~aal,
    title = "AAL, 2020-2050"
  )
```
<br> Here we can visually see the AAL across our block groups. The darkest chunk has the middle-most amount of buildings but the greatest loss, the orange has the most buildings and a little less loss, the white chunk has the least amount of buildings and the least amount of loss. This probably has to do with the fact that the white chunk is the most inland versus the darkest chunk which is closest to the water and the canals.Waterfront properties generally are more expensive than inland ones so it would makes sense that despite there being a fair amount of buildings, there is a huge loss compared to the middle chunk. Even at the block level, however, our results may not be granular enough to fully understand the placement of the buildings and their associated AALs. This would be an interesting starting point to look at cost of housing, housing tenancy, and AAL affects.  <br> 
