---
title: "Week8_ChapterWalkthrough"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(censusapi)
library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(leaflet)
library(StatMatch)


Sys.setenv(CENSUS_KEY="c18a7f6e5cbbd1e7f0d3047d8b674cbaad89bcb3")
```
Match Simmilarity Across Census Tracts
```{r}
acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_tracts <-
  tracts("CA", bay_county_names, cb = T, progress_bar = F)

bay_multiple_tract <- 
  getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "tract:*",
    regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
    vars = c(
      "B06009_001E",
      "B06009_002E",
      "B06009_003E",
      "B19001_001E",
      "B19001_014E",
      "B19001_015E",
      "B19001_016E",
      "B19001_017E",
      "B19001A_001E"
    )
  ) %>% 
  transmute(
    tract = paste0(state, county, tract),
    perc_college = 1 - (B06009_002E + B06009_003E) / B06009_001E,
    perc_over100k = (B19001_014E + B19001_015E + B19001_016E + B19001_017E) / B19001_001E,
    perc_white = B19001A_001E / B19001_001E
  ) %>% 
  filter(
    !is.na(perc_college), 
    !is.na(perc_over100k),
    !is.na(perc_white)
  )

#Does algebra, all numbers
obs_matrix <-
  bay_multiple_tract %>% 
  select(
    perc_white, 
    perc_over100k,
    perc_college
  ) %>% 
  as.matrix()

##Distance matrix, requires the StatMatch package
dist_matrix <- mahalanobis.dist(obs_matrix)

##add row names and column names to all rows/columns
rownames(dist_matrix) <- bay_multiple_tract$tract
colnames(dist_matrix) <- bay_multiple_tract$tract

##Make a copy matrix to pair each tract to its twin  
dist_matrix_pairmatch <- dist_matrix
diag(dist_matrix_pairmatch) <- NA

##data frame with the matched tracts
matched_pair_tract <-
  1:nrow(dist_matrix_pairmatch) %>% 
  map_dfr(function(x){
    
    min_index <- which(dist_matrix_pairmatch[x, ] == min(dist_matrix_pairmatch[x, ], na.rm = T)) ##true or false question 
   ##build data frame 
    data.frame(
      tract = bay_multiple_tract$tract[x],
      matched_tract = bay_multiple_tract$tract[min_index]
    )
  })
```

Map

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[2,1])
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[2,2])
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[3,1]),
    color = "green"
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[3,2]),
    color = "green"
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[4,1]),
    color = "red"
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[4,2]),
    color = "red"
  )
```

Map A Particular Tract via Difference in Difference Concept
```{r}
match_set_tract <- dist_matrix["06081611900", ] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    tract = rowname,
    match = "."
  ) %>% 
  right_join(
    bay_multiple_tract
  ) %>% 
  arrange(match) %>% 
  .[1:21, ] %>% 
  left_join(bay_tracts %>% select(tract = GEOID)) %>% 
  st_as_sf()

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match_set_tract[1, ],
    color = "red"
  ) %>% 
  addPolygons(
    data = match_set_tract[-1, ]
  )
```


CHAPTER 6.2 Difference in Difference Technique

Generate PUMAS

```{r}
ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_pumas,
    weight = 1,
    color = "gray",
    label = ~PUMACE10
  ) %>% 
  addMarkers(
    lng = -121.9415017, ##directly add long and lat. 
    lat = 37.502171
  ) %>% 
  addPolygons(
    data = bay_pumas %>% 
      filter(PUMACE10 == "08504")
  )
```

```{r}
pums_2014_2019 <- readRDS("pums_2014_2019_wts.rds")

pums_bart <- pums_2014_2019 %>%
  mutate(
    PWGTP = as.numeric(PWGTP),
    bart = ifelse(
      JWTR %in% c("4"),
      PWGTP,
      0
    )
  ) %>% 
  ##BART Ridership
  group_by(PUMA, year) %>% 
  summarize(
    pop = sum(PWGTP),
    bart = sum(bart)
  )

```
Map distribution of population and BART commuters in the Bay Area PUMAs

```{r}
pums_pal <- colorNumeric(
  palette = "YlOrRd",
  domain = pums_bart %>% 
    filter(year == 2017) %>% 
    pull(pop)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2017) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(pop),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": Population ", pop)
  )
```
Livermore and Pleasonton have a lot of people in it, but the PUMA is also very very big compared to the SF PUMA. This raw count is very misleading because of that. Oakland and Emeryville have high users based on its acerage. 


Map Bart Users
```{r}
pums_pal <- colorNumeric(
  palette = "GnBu",
  domain = pums_bart %>% 
    filter(year == 2017) %>% 
    pull(bart)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2017) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(bart),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": ", bart, " BART commute riders")
  )
```
Some of the small PUMAs in Berkely, SF, and Oakland seem to have a lot of BART riders. But again this could be skewed becuase of its size. Ie, Livermore has a sparse number of BART riders in the area but it also has a very high acerage. 


Clean Data by removing pop data

```{r}
pums_bart_clean <-
  pums_bart %>% 
  select(-pop) %>% 
  pivot_wider( 
    names_from = year,
    values_from = bart
  )
```


Matching Technique

```{r}
obs_matrix <-
  pums_bart_clean %>% 
  ungroup() %>% 
  select(`2014`,`2015`,`2016`) %>% ##add 2017 for the pset
  as.matrix()

dist_matrix <- mahalanobis.dist(obs_matrix)

rownames(dist_matrix) <- pums_bart_clean$PUMA
colnames(dist_matrix) <- pums_bart_clean$PUMA

match <- dist_matrix["08504",] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    PUMA = rowname,
    match = "."
  ) %>% 
  right_join(
    pums_bart_clean
  ) %>% 
  arrange(match) %>% 
  .[1:11, ] %>% 
  left_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
  st_as_sf()
```
Map the Matrix Data

```{r}
leaflet() %>% 
  addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match[1, ],
    color = "red",
    label = ~PUMA
  ) %>% 
  addPolygons(
    data = match[-1, ],
    label = ~PUMA
  )
```

Difference in Differnece Treatment

```{r}
match_pumas <-
  match %>% 
  filter(!PUMA %in% c("08504")) %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>%
  group_by(
    year
  ) %>% 
  summarize(
    bart = mean(bart),
    PUMA = "Similar PUMAs"
  )

treatment_pumas <-
  match %>% 
  filter(PUMA %in% c("08504")) %>% 
  select(-match) %>% 
  st_drop_geometry() %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  )

rbind(
  treatment_pumas,
  match_pumas
) %>% 
  ggplot(
    aes(
      x = as.numeric(year),
      y = bart,
      color = PUMA
    )
  ) +
  geom_line() +
  geom_vline(xintercept = 2017, linetype = "dashed") +
  labs(
    title = "Milpitas vs. control neighborhoods, BART ridership",
    x = "Year",
    y = "BART commute riders"
  )
```
Diff in Diff Regression

```{r}
transit_did <-
  match %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>% 
  mutate(
    year = year %>% as.numeric(),
    time = ifelse(year >= 2017, 1, 0),
    treated = ifelse(PUMA == "08504", 1, 0)
  )

did_reg <- lm(bart ~ treated*time, data = transit_did)

summary(did_reg)
```

